---
title: "group assignment"
author: "sammy, Joseph, Chiomma"
date: "2025-02-14"
output:  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# GROUP EXCELLENCE

### Knowing my current working directory

```{r}
getwd()

```
### Setting up seed to avoid the reassignment of random numbers so we get a consistent result after every successful exercution of code.
```{r}
set.seed(255)

```

###uploading csv file as dd
```{r}
attach(Maternal_Health_Risk_Data_Set_Copy)
```
### Getting to know the kind of data we are working with.
```{r}
summary(Maternal_Health_Risk_Data_Set_Copy)

```

### Assigning the name _dd_ to our dataset so we can easily call them during coding
```{r}
dd <- Maternal_Health_Risk_Data_Set_Copy
dd

```
```{r}
summary(dd)
```


##(i)	Find the mean age of each risk level and arranging the results in descending order

```{r}
q <- dd %>% group_by(RiskLevel) %>% summarise(mean_age = mean(Age)) 

```
### Then we use the arranging in descending order code.
```{r}
#q_in_desc <- q %>% arrange(desc(mean_age))
```


###Using the pipe to get a straight result, we have;

```{r}
dd %>% group_by(RiskLevel) %>% summarise(mean_age = mean(Age)) %>% arrange(desc(mean_age))
```


##(ii)	Find the median diastolic of each risk level and order them in ascending order. 


```{r}
dd %>% group_by(RiskLevel) %>% summarise(median_DiastolicBP = median(DiastolicBP) ) %>% arrange(median_DiastolicBP)


```
and order them in ascending order. 


```{r}
#dd %>% group_by([,cRiskLevel) %>% summarise(median_DiastolicBP = median() )
```
```{r}

```
## 2.	Create a new variable name it systolic rate. Where systolic rate = systolic BP/age

```{r}
dd_1 <- mutate(dd, systolic_rate = SystolicBP/Age)
dd_1
```

#  3.	Filter participants of age <= 40 years, determine average systolic rate for each risk level

```{r}
dd_1 %>% filter(Age <= 40, ) %>% group_by(RiskLevel) %>% 
  summarise(avg_systolic = mean(systolic_rate))

```

## 4.	For each continuous variable in the dataset, create histograms and box plots, and provide insights on the data distribution. Additionally, calculate the five-number summary for each variable, grouped by risk level.

# Question 4b. Summary Statistics
```{r}
# summary grouped by risk level
 dd %>%
  group_by(RiskLevel) %>% 
  summarise(across(all_of(dd), 
                   list(Min = min, 
                        Q1 = ~quantile(.x, 0.25), 
                        Median = median, 
                        Q3 = ~quantile(.x, 0.75), 
                        Max = max), 
                   .names = "{.col}_{.fn}"))
## how can we rearrange the table
```


```{r}
names(dd)
```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

```


```{r}
#hist(dd$Age)
```
```{r}
hist(dd$SystolicBP)
```
```{r}
hist(dd$DiastolicBP)
```
```{r}
hist(dd$BS)
```

```{r}
hist(dd$BodyTemp)
```
```{r}
hist(dd$HeartRate)
```
# box plot
```{r}
boxplot(HeartRate~BodyTemp)
```
```{r}
boxplot(HeartRate~BS)
```
```{r}
boxplot(DiastolicBP~BS)
```
```{r}
boxplot(HeartRate~BodyTemp)
```




##5.	A) Explore the relationship between heart rate and the following variables using suitable graphical methods:
(i) Age
(ii) Systolic blood pressure
(iii) Diastolic blood pressure
(iv) Body temperature
Provide comments on your observations.


## Explore the relationship between heart rate and the following variables using suitable graphical methods

```{r}
plot(Age, HeartRate)
```
```{r}
plot(HeartRate~SystolicBP)
```

```{r}
plot(HeartRate~DiastolicBP)
```
```{r}
plot(HeartRate~BodyTemp)
```
##B) Based on the results from A) fit a simple linear model using heart rate as a response variable and the variables in A as predictor variables. Interpret your results.


```{r}
MODEL1<-lm(HeartRate~Age)
summary(MODEL1)
```


```{r}
MODEL2<-lm(HeartRate~SystolicBP)
summary(MODEL2)
```
```{r}
MODEL3<-lm(HeartRate~DiastolicBP)
summary(MODEL3)
```
```{r}
MODEL4<-lm(HeartRate~BodyTemp)
summary(MODEL4)
```

```{r}
MODEL5<-lm(HeartRate~Age+SystolicBP+DiastolicBP+BodyTemp  )
summary(MODEL5)
``` 

```{r}
names(dd)
```


```{r}

fitall<-lm(HeartRate~ Age + SystolicBP + DiastolicBP + BodyTemp ,data=dd)
summary(fitall)
```

```{r}
fitstart<-lm(HeartRate~1, data=dd)
summary(fitstart)
```
```{r}
stepwisemodel<-step(fitall,direction = "both")
summary(stepwisemodel)
```





##Residual analysis


```{r}
install.packages("car")
install.packages("carData ")
library(car)
library(carData)


```{r}
vif(residual)
```

```{r}
residual<-stepwisemodel$residuals
plot(dd$Age,residual)
```
```{r}
residual<-stepwisemodel$residuals
plot(dd$SystolicBP,residual)
```

```{r}
residual<-stepwisemodel$residuals
plot(dd$DiastolicBP,residual)
```

```{r}
residual<-stepwisemodel$residuals
plot(dd$BodyTemp,residual)
```

```{r}
plot(stepwisemodel,which=1,col=c("blue"))
```




```{r}
plot(residual)
```


```{r}
t.test(residual)
```
Ho:residual mean is equal to zero
H1:residual mean is not equal to zero.
since our p-value of 1 is greater than our alpha value of 0.05, we fail to reject the null hypothesis and conclude that the residual mean is equal to zero. 


```{r}
library(lmtest)
```

###Test of constant variance

```{r}
plot(residual)
```
from the residual plot above,it appears the residuals are homoscedastic, but we need to confirm this by conducting the BRUESCH PAGAN test. 

```{r}
bptest(stepwisemodel)
```
Ho: homoscedastic
H1:heteroscedastic
since our p-value(0.04686) is less than an alpha value of 0.05,we reject the null hypothesis and conclude that there is heteroscedasticity.


##Residual test of independence
```{r}
dwtest(stepwisemodel)
```
Ho:there is no residual autocorrelation 
H1:there is residual autocorrelation
since our p-value(0.9476) is greater than an alpha value of 0.05,we fail to reject the null hypothesis and conclude that there is no autocorrelation among the residuals.
we will however proceed to plot the autocorrelation function (acf) to confirm.


```{r}
acf(residual)
```

## test of normality of residuals.

```{r}
hist(residual)
```
from the histogram, it is not show clear whether the residuals are normally distributed or not, so we proceed to conduct the shapiro test of normality.

```{r}
shapiro.test(residual)
```

Ho:residuals are normally distributed
H1: residuals are not normally distributed

since the p-value (2.2e^-16) is less the (alpha=0.05),we reject the Ho and conclude that the residuals are not normally distributed. 

























